---
title: "Statistical Inference Project 1"
author: "Vijay Shankar"
date: "14 November 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Statistical Inference Project Part 1: Simulation Exercise Instructions

The objective of this project is to investigate the exponential distribution in R and compare it with the Central Limit Theorem. The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also 1/lambda. Set lambda = 0.2 for all of the simulations. The investigation wuld be of the distribution of averages of 40 exponentials. The project requires thousand simulations.

```{r, echo=TRUE, message=FALSE, warning=FALSE}

set.seed(100)

# Given:
lambda <- 0.2

n <- 40

sims <- 1000

# simulate
sims_exponentials <- replicate(sims, rexp(n, lambda))

# calculate mean of exponentials
means_exponentials <- apply(sims_exponentials, 2, mean)

```

Illustrate via simulation and associated explanatory text the properties of the distribution of the mean of 40 exponentials. 

### Question 1: Show the sample mean and compare it to the theoretical mean of the distribution.
The theoretical mean of the distribution can be calculated by calculating 1/lambda. The sample mean can be calculated by using the mean function over the means_exponentials taht we generated above.

#### Sample Mean

```{r, echo=TRUE, message=FALSE, warning=FALSE}

sample_mean <- mean(means_exponentials)

sample_mean
```

#### Theoretical Mean

```{r, echo=TRUE, message=FALSE, warning=FALSE}

Theoretical_mean <- 1/lambda
Theoretical_mean

```

#### Showing the comparison

```{r, echo=TRUE, message=FALSE, warning=FALSE}

hist(means_exponentials, xlab = "Mean", main = "Illustration of Means")
abline(v = sample_mean, col = "blue")
abline(v = Theoretical_mean, col = "red")

```

#### Question 1 conclusion
The sample mean and the theoretical mean are very close as shown the the graph.

### Question 2 : Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.

The variance of the sample can be obtained by squaring the standard deviation of the sample that can be obtained using the sd function. 

The standard deviation calculated theoretically is (1/lambda)/squareroot(n). The variance can be obtained by squaring this figure.

```{r, echo=TRUE, message=FALSE, warning=FALSE}

sd_sample <- sd(means_exponentials)

var_sample <- (sd_sample^2)

var_sample

sd_theoretical <- (1/lambda)/(sqrt(n))

var_theoretical <- (sd_theoretical^2)

var_theoretical

# Comparison

abs(sd_sample-sd_theoretical)

```

#### Question 2 Conclusion

The variance calculated for the sample and that calculated using theory are very close (varying by 0.01145566)

###Question 3: Show that the distribution is approximately normal.

In point 3, focus on the difference between the distribution of a large collection of random exponentials and the distribution of a large collection of averages of 40 exponentials.

In order to show that the distribution is approximtely normal, the density histogram of the sample should be overlayed with the normal distribution with mean of 1/lambda and standard deviation of (1/lambda/sqrt(n)).

```{r, echo=TRUE, message=FALSE, warning=FALSE}

xfit <- seq(min(means_exponentials), max(means_exponentials), length=100)
yfit <- dnorm(xfit, mean=1/lambda, sd=(1/lambda/sqrt(n)))
hist(means_exponentials,breaks=n,prob=T,col="light blue",xlab = "Means",main="Density of means",ylab="Density")
lines(xfit, yfit, pch=22, col="black", lty=5)

```

The distribution the plot looks close to normal distribution. However, a better way of determining this, a better approach would be to plot the quantiles of the normal distribution and quantiles of the means of exponentials in a Q-Q plot.

```{r, echo=TRUE, message=FALSE, warning=FALSE}

qqnorm(means_exponentials, col=2)
qqline(means_exponentials, col = 1)

```

#### Question 3 Conclusion
Based on the histogram and the Q-Q plot, we can say that the distribution is approximately normal

### Assumptions
In this exercise we used very high value for the simulations. Had we used a much lower number, say 20 instead of 1000, the values would have been different. This is because assymptotics are a limit where sample size tends towads infinity.

This can be shown by the code chunk below:

```{r, echo=TRUE, message=FALSE, warning=FALSE}
lambda <- 0.2

n <- 40

sims <- 20

sims_exponentials <- replicate(sims, rexp(n, lambda))

means_exponentials <- apply(sims_exponentials, 2, mean)


sample_mean <- mean(means_exponentials)

sample_mean


Theoretical_mean <- 1/lambda
Theoretical_mean


hist(means_exponentials, xlab = "Mean", main = "Illustration of Means")
abline(v = sample_mean, col = "blue")
abline(v = Theoretical_mean, col = "red")


sd_sample <- sd(means_exponentials)

var_sample <- (sd_sample^2)

var_sample

sd_theoretical <- (1/lambda)/(sqrt(n))

var_theoretical <- (sd_theoretical^2)

var_theoretical

# Comparison

abs(sd_sample-sd_theoretical)


xfit <- seq(min(means_exponentials), max(means_exponentials), length=100)
yfit <- dnorm(xfit, mean=1/lambda, sd=(1/lambda/sqrt(n)))
hist(means_exponentials,breaks=n,prob=T,col="light blue",xlab = "Means",main="Density of means",ylab="Density")
lines(xfit, yfit, pch=22, col="black", lty=5)


qqnorm(means_exponentials, col=2)
qqline(means_exponentials, col = 1)

```



