---
title: "Relationship Between Transmission Type & Mileage"
author: "Motor Trend - Vijay Shankar"
date: "December 15, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary
Motor Trend magazine are interested in the following aspects:
1. Is an automatic or manual transmission better for MPG
2. Quantify the MPG difference between automatic and manual transmissions
## Initial Exploratory Analysis and Data Transformation
An initial exploratory analysis of the data set is performed and submitted below. Plots / additional information is submitted in Appendix A of this report. 
```{r, echo=FALSE, message=FALSE, warning=FALSE,results='hide'}
library(ggplot2)
data("mtcars")
head(mtcars)
str(mtcars)

## Data Transformation ofr factor variables that are shown as number variables in the dataset
mtcars$cyl  <- factor(mtcars$cyl)
mtcars$vs   <- factor(mtcars$vs)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
mtcars$am   <- factor(mtcars$am,labels=c("Automatic","Manual"))

```
Further, it can be gleaned from Apendix A (Figure 1) that there was an impact on MPG by transmission with Automatic transmissions having a lower MPG.
## Regression Analysis
### Model Selection
Let us build an initial model with only am as the regressor and call this fit1. We will then include the above listed variables (including am) as fit2 and fit 3 would include all the variables.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
fit1 <- lm(mpg ~ am, data = mtcars)
summary(fit1)
fit2 <- lm(mpg~am + cyl + disp + hp + drat + wt + vs, data = mtcars)
fit3 <- lm(mpg ~. , data = mtcars)
```
It can be concluded here that average MPG for automatic is 17.1 MPG, while manual is 7.2 MPG higher.Lets compare these models:
```{r, echo=TRUE, message=FALSE, warning=FALSE}
anova(fit1,fit2,fit3)
summary(fit2)
```
It can be seen from the summary of fit 1 above that the model with only am explains approximately 36% of the data (R^2). The anova function shows that owing to the p-value, inclusion of other regressors adds value to the analysis. Fit 3 does not add value to the model. The model fit2 explains approximately 87% of the data. From fit 2 we can show that difference between automatic and manual transmissions is 2.56 MPG taking other regressors into consideration.
## Residuals
Figure 3 in Apendix A studies the residuals to understand the liverage points for fit2.
## Inference
Assuming that the transmission data has a normal distribution, let us perform a t-test to see if manual and auto transmissions are clearly different.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
t.test(mpg ~ am, data = mtcars)
```
## Conclusion
Based on the observations from our model, we can conclude the following,
1. Cars with Manual transmission get more miles per gallon mpg compared to cars with Automatic transmission. (2.56 adjusted by cyl, disp, hp, drat, wt and vs).
2. mpg will decrease by 2.6 (adjusted by cyl, disp, hp, drat,vs and am) for every 1000 lb increase in wt.
3. mpg decreases negligibly with increase of hp.
4. If number of cylinders, cyl increases from 4 to 6 and 8, mpg will decrease by a factor of 2 and 0.02 respectively (adjusted by wt, disp, hp, drat,vs and am). 

## Appendix A
```{r, echo=TRUE, message=FALSE, warning=FALSE}

boxplot(mpg ~ am, data = mtcars, col = (c("indianred","turquoise")), ylab = "Mileage (MPG)", xlab = "Transmission Type")
```
Figure 1 Plotting Mileage by Transmission Type 

It was concluded that the variables; cyl, disp, hp, drat, wt, vs and am have a strong correlation with MPG.
```{r, echo=TRUE, message=FALSE, warning=FALSE}

pairs(mpg ~ ., data = mtcars)

```
Figure 2 Pairs Plot for Understanding Correlations

```{r, echo=TRUE, message=FALSE, warning=FALSE}
par(mfrow = c(2,2))
plot(fit2)
```
Figure 3 Residuals & Diagnostics
From the Figure 3, we can make the following observations,
1. The points in the Residuals vs. Fitted plot seem to be randomly scattered on the plot and verify the independence condition.
2. The Normal Q-Q plot consists of the points which mostly fall on the line indicating that the residuals are normally distributed.
3. The Scale-Location plot consists of points scattered in a constant band pattern, indicating constant variance.
4. There are some distinct points of interest (outliers or leverage points) in the top right of the plots.
